---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# WASH household Survey (2018 - 2023)

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

<!-- badges: end -->

This dataset contains results from the Sustainable Development Goals
(SDG) Household-Level Survey for WASH conducted by BASEFlow Malawi between 2018
and 2023, using the mWater platform. The data provides detailed,
household-level insights into access to water, sanitation, hygiene, and
environmental conditions across different communities in Malawi.

Each row represents a completed household survey and includes:

-   Survey metadata: submission date, consent, GPS location

-   Water point information: type, method of collection, region,
    location accuracy, altitude

-   Water quality indicators: microbial tests, confidence levels,
    associated health risks

-   Sanitation: toilet type, waste removal and treatment methods

-   Hygiene: handwashing facility availability, presence of water and
    soap

-   Other socio-environmental factors: including responses to structured
    and open-ended questions

**🔍 Use Cases**

This dataset is suitable for a range of applications, including:

1.  WASH Program Monitoring & Evaluation Evaluate household access to
    clean water and sanitation services over time and across regions.

2.  Public Health Research Identify environmental and behavioral factors
    contributing to hygiene-related illnesses.

3.  Geospatial Analysis Map infrastructure such as boreholes and
    sanitation facilities to analyze regional disparities.

4.  Evidence-Based Planning & Advocacy Support national and local
    policymakers in achieving SDG 6: Clean Water and Sanitation for All.

5.  Education and Capacity Building Use in training programs and
    university courses on data-driven development, environmental health,
    or digital health systems.

6.  Data Science and ML Train models to predict risk patterns or
    infrastructure needs based on survey indicators.

## Installation

You can install the development version of washmalawi from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/washmalawi")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in your
    browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/washmalawi/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to Sustainable Development Goals (SDG) Household-Level Survey data for Malawi between 2018 and 2023

```{r}
library(washmalawi)
```

### washmalawi

The dataset `washmalawi` contains
`r nrow(washmalawi)` observations and `r ncol(washmalawi)`
variables

```{r}
washmalawi |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "washmalawi.rda") |>
  dplyr::select(variable_name:description) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

## Example

```{r}
# Plot a bar chart for the type of toilets in use
library(washmalawi)
library(dplyr)
library(ggplot2)
library(maps)
# Filter out records with missing or empty 'toilet_type'
# Then count the number of occurrences for each toilet type
toilet_data <- washmalawi %>%
  filter(!is.na(toilet_facility_type) & toilet_facility_type != "NA") %>%
  count(toilet_facility_type, sort = TRUE)

# Create a horizontal bar chart showing the number of households
# using each type of toilet
ggplot(toilet_data, aes(x = reorder(toilet_facility_type, n), y = n)) +
  geom_bar(stat = "identity", fill = "#2C7BB6") +  # Use blue bars for counts
  coord_flip() +  # Flip axes to make the chart horizontal
  labs(
    title = "Distribution of Toilet Types in Surveyed Households",  # Chart title
    x = "Toilet Type",  # X-axis label showing toilet types)
    y = "Number of Households"  # Y-axis label (number of entries)
  ) +
  theme_minimal(base_size = 13)


# Plots maps for the water collection points 
# Filter usable water point locations
water_map_data <- washmalawi %>%
  filter(!is.na(latitude) & !is.na(longitude))  # Ensure geo-points are valid

# Get Malawi map outline
malawi_map <- map_data("world", region = "Malawi")

# Plot the water point locations
ggplot() +
  geom_polygon(data = malawi_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray70") +
  geom_point(data = water_map_data,
             aes(x = longitude, y = latitude, color = water_source_type),
             size = 2, alpha = 0.7) +
  coord_fixed(1.3) +
  labs(
    title = "Geo-Referenced Water Point Locations in Malawi",
    x = "Longitude", y = "Latitude", color = "Water Source Type"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/%7B%7B%7Bpackagename%7D%7D%7D/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("washmalawi")
```
