---
title: "sdghhsurveywm: Package review"
author: "Colin Walder"
format: html
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(DataExplorer)
package_data <- read_csv("sdghhsurveymw.csv")
```

# General Information and Metadata

-   Package Name: Too long of an acronym and it seems to be more about WASH than SDGs. What about washmalawi? Please suggest a few names. **RESOLVED**
-   The Title and Description fields are missing in DESCRIPTION. **RESOLVED**

# Data Content & Quality

```{r, eval=FALSE}
# create_report is the main function from the DataExplorer package (see https://boxuancui.github.io/DataExplorer/)
# run it to get a quick overview of your data
create_report(package_data)
```

-   What is the reason behind including non-consenting respondents? **RESOLVED**

```{r}
package_data |> 
  count(survey_consent) |> 
  ggplot(aes(x = survey_consent, y = n, label = n)) +
  geom_col() +
  geom_label()
```

-   The variables `link_community_site` and `link_community_na` are not documented in the dictionary. **RESOLVED**
-   `w1_method`: Only "GPS" in the dataset but you also mention interviews in the variable description. **RESOLVED**

```{r}
count(package_data, w1_method)
```

-   `w1_region`: I'd delete Malawi as all observations are from Malawi. In addition, split up traditional authority and district into two variables for easier data analysis later. **RESOLVED**

-   Explain in more detail how `w1_latitude` is different from `w1_latitude_loc` in the dictionary. **RESOLVED**

-   Remove the numbers/abbreviations and spell out the meaning of all variables, e.g., `water_point_latitude` rather than `w1_latitude`. **RESOLVED**

-   For the variables below, all values are NA. Either add values to these columns or delete the columns as they don't add any information for now.

```{r}
find_all_na_columns <- function(df) {
  names(df)[colSums(is.na(df)) == nrow(df)]
}

find_all_na_columns(package_data)
```

-   `w5` seems to be heavily skewed. Is this actually the case or might there be some data entry errors? **RESOLVED**

```{r}
hist(package_data$w5)

quantile(package_data$w5, probs = seq(0,1,0.1), na.rm = TRUE)
```

# Data Processing

-   Why are there some many rows with NAs in latitude? **RESOLVED**
-   Remove `check_utf8()`. This is not part of the data processing, also considering that there are no non-UTF-8 characters found in the dataset when running the function. **RESOLVED**
-   As a result, also delete the conversion from Latin1 to UTF-8 (lines 57-67). **RESOLVED**
-   Delete lines starting from line 91 to the end as data visualization is not part of the data processing. **RESOLVED**

# Documentation

-   The title is cut off in the visualization.
-   Add a second visualization with a map to show the users that the data is geo-referenced.
